NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-03-21 Manuscript/manuscriptPS.tex
Add sodium binding results to DMPS/DMPC bilayers

Since systems with negatively charged PS lipids always contain counterions
and the ion binding presumably affects the PS headgroup
order parameters, the discussion of ion binding affinity and
headgroup structure cannot be separated as done previously
for PC lipids \cite{botan15,catte16}. The density profiles
of counterions along membrane normal from different simulations
are shown in Fig. \ref{NAdensPOPS}.
Significant differences are observed in counterion
densities between different simulation models. 
\todo{We should try to figure out which one of these are more
  realistic. One option could be to use PS headgroup order parameter
  data measured as a function of NaCl \cite{roux86}. Another
  approach could be to use the area per molecule, which is suggested to
  depend on the binding affinity of counterions \cite{pandit02,mukhopadhyay04,pedersen06}. The
  experimental value for area per molecule is available at \cite{pan14}.}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

>>>>

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Since systems with negatively charged PS lipids always contain counterions
and the ion binding presumably affects the PS headgroup
order parameters, the discussion of ion binding affinity and
headgroup structure cannot be separated as done previously
for PC lipids \cite{botan15,catte16}. The density profiles
of counterions along membrane normal from different simulations
show significant binding of counterions in bilayers in all
models in Fig. \ref{NAdensPOPS}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}
Differences are, however,
observed in the total binding affinity and in the shape of density profiles.
Comparison of bulk concentrations suggests that strongest binding
is observed in MacRog, Berger and Lipid17 models, respectively.
Other models gives roughly similar total binding affinity
based on rough comparison of bulk concentrations.
In CHARMM36, CHARMM36ua and Gromos-CKP models two maxima are observed
in sodium density plots, possible corresponding binding to phoshate
and carboxylic grops in the PS headgroup. In other models only
a single maxima is observed. \todo{More quantitative analysis of
binding positions and affinity may be useful.}

Based on previous simulation results it has been suggested that
the counterion binding and concomitant screening of headgroup repulsion
leads to smaller area per lipid in PS lipid bilayers than in PC
bilayers \cite{pandit02,mukhopadhyay04,pedersen06}. Therefore we also show the area per lipid from different models
in Fig. \ref{NAdensPOPS}. The proposed trend is observed for Lipid17 model,
for which the Joung-Cheatham ions give higher affinity and smaller area
per molecule. However, the trend is not observed when comparing accross different
force fields. For example, CHARMM36ua and MacRog give similar area per molecule
but binding affinity is significantly higher in MacRog. On the other hand,
Gromos-CKP models gives a larger area per lipid than other models, being only
in agreement with experiments \cite{pan14}, but the binding affinity is comparable
to the CHARMM36, CHARMM36us and Slipids, which give smaller are per molecule.

In order to compare the sodium interaction with PS containing membranes between
experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the order parameters of the
$\alpha$ carbon of PS headgroup measured from the mixture with PC as a
function of added NaCl \cite{roux86} together with the resuls from CHARMM36 model. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
  \caption{\label{PSresponseTONaCl}
    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
  }
\end{figure}
The experimental results show essentially no changes in the order parameters as a function of
added NaCl, while significant changes are observed in simulations. However,
the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
Therefore, we cannot exclude the possibility that the NaCl induced changes were already
saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
\todo{It would be probably better to use datasets for NaCl and KCl from Ref. \citenum{roux90}
to compare monovalent ion binding to PS containing bilayer.}
